Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis

### 2018-04-11 jhir/latex-manuscript/manuscript.tex
finalized manuscript layout

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{geometry}
\usepackage{pdflscape}
%
\usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
\usepackage{rotating}
% Insert the name of "your journal" with
\journalname{J Healthc Inform Res}
%
\begin{document}

\title{Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis
}
%\subtitle{Do you have a subtitle?\\ If so, write it here}

%\titlerunning{Short form of title}        % if too long for running head

\author{Mehedi Hasan, BS$^1$\and 
April Idalski Carcone, PhD$^2$\and 
Sylvie Naar, PhD$^3$\and 
Susan Eggly, PhD$^4$\and  
Gwen L. Alexander, PhD$^5$\and 
Kathryn E Brogan Hartlieb, PhD$^6$\and 
Alexander Kotov, PhD$^1$
}

%\authorrunning{Short form of author list} % if too long for running head

\institute{Alexander Kotov (corresponding author) \at
              5057 Woodward Suite 14001.6 \\
              Detroit, MI 48202\\
              Tel.: +1(313) 577-9307\\
              \email{kotov@wayne.edu} 
           \and
           $^1$Department of Computer Science, College of Engineering, Wayne State University, Detroit, MI 48202\\
$^2$Division of Behavioral Health Sciences, Department of Family Medicine and Public Health Sciences, Wayne State University School of Medicine, Detroit, MI 48202\\
$^3$Director, Center for Translational Behavioral Research, Department of Behavioral Sciences and Social Medicine, Florida State University, FL 32306\\  
$^4$Department of Oncology, Wayne State University/Karmanos Cancer Institute, Detroit, MI 48201\\
$^5$Department of Public Health Sciences, Henry Ford Health System, Detroit, MI 48202\\
$^6$Department of Humanities, Health and Society, Wertheim College of Medicine, Florida International University, Miami, FL 33199    
}

@@

\keywords{sequential analysis \and hidden markov model \and closed frequent pattern mining \and motivational interviewing \and weight loss \and adolescents}
%\PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{sec:intro}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change \citep{miller2013motivational}. The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills \citep{miller2013motivational}. The relational hypothesis suggests that counselors’ use of accurate empathy, positive regard, and congruence create the “spirit of MI”, an optimal therapeutic state to explore behavior change. MI’s technical hypothesis \citep{miller2009toward} states that counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient “change talk”. Change talk are patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change. Previous studies [3] have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention [4]. In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk). Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk [5]. These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers’ ability to test MI’s technical hypothesis.

@@

% For tables use
\begin{table}
% table caption is above the table
\caption{MYSCOPE codebook.}
\label{tab:codebook}       % Give a unique label
% For LaTeX tables use
\begin{tabular}{lll}
\hline\noalign{\smallskip}
first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number \\
number & number & number \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

@@

% For two-column wide figures use
\begin{figure*}
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[width=1.0\textwidth]{figures/bic-vs-hidden-states.eps}
% figure caption is below the figure
\caption{Bayesian information criterion (BIC) of HMM models of successful (left) and unsuccessful (right) interviews by varying the number of hidden states.}
\label{fig:bic}       % Give a unique label
\end{figure*}

@@

% For two-column wide figures use
\begin{figure*}
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[width=0.95\textwidth]{figures/fpm.eps}
% figure caption is below the figure
\caption{Different type of itemsets is illustrated with a frequent pattern mining method applied on transaction database by considering minimum support count 2, where TID represents the transaction ID of the corresponding items.}
\label{fig:fpm}       % Give a unique label
\end{figure*}

@@

%\newgeometry{margin=2cm} % modify this if you need even more space
%\newgeometry{left=4cm}
\begin{sidewaystable}
\caption{Hidden Markov Model Emission Matrices}
\label{tab:emission}       % Give a unique label
% For LaTeX tables use
\begin{tabular}{lll}
\hline\noalign{\smallskip}
first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number \\
number & number & number \\
\noalign{\smallskip}\hline
\end{tabular}

@@

\caption{Hidden Markov Model Transition Matrices}
\label{tab:transition}       % Give a unique label
% For LaTeX tables use
\begin{tabular}{lll}
\hline\noalign{\smallskip}
first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number \\
number & number & number \\
\noalign{\smallskip}\hline
\end{tabular}
\end{sidewaystable}

%\restoregeometry

@@

% For tables use
\begin{table}
% table caption is above the table
\caption{Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences.}
\label{tab:patterns}       % Give a unique label
% For LaTeX tables use
\begin{tabular}{lll}
\hline\noalign{\smallskip}
first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number \\
number & number & number \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

@@

%Text with citations \citep{wei2006semi} and \citep{keogh2000scaling}.

%\paragraph{Paragraph headings} Use paragraph headings as needed.
%\begin{equation}
%a^2+b^2=c^2
%\end{equation}

@@

% BibTeX users please use one of
%\bibliographystyle{spbasic}      % basic style, author-year citations
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
\bibliographystyle{natbib}       % JHIR format
%\bibliographystyle{spphys}       % APS-like style for physics
\bibliography{references}   % name your BibTeX data base

\end{document}
% end of file template.tex

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{geometry}
\usepackage{pdflscape}
%
\usepackage{mathptmx}      % use Times fonts if available on your TeX system

\usepackage{rotating}
\usepackage{tabularx}
\usepackage[flushleft]{threeparttable}

% Insert the name of "your journal" with
\journalname{J Healthc Inform Res}
%
\begin{document}

\title{Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis
}

\author{Mehedi Hasan$^1$\and 
April Idalski Carcone$^2$\and 
Sylvie Naar$^3$\and 
Susan Eggly$^4$\and  
Gwen L. Alexander$^5$\and 
Kathryn E Brogan Hartlieb$^6$\and 
Alexander Kotov$^1$
}

\institute{Mehedi Hasan \at
              \email{mehedi@wayne.edu} 
           \and
           April Idalski Carcone \at
              \email{acarcone@med.wayne.edu} 
           \and
           Sylvie Naar \at
              \email{sylvie.naar@med.fsu.edu}
           \and
           Susan Eggly \at
              \email{egglys@karmanos.org} 
           \and
           Gwen L. Alexander \at
              \email{GALEXAN2@hfhs.org}
           \and
           Kathryn E Brogan Hartlieb \at
              \email{hartlieb@fiu.edu} 
           \and
           Alexander Kotov (corresponding author) \at
              Tel.: +1(313) 577-9307 \\
              \email{kotov@wayne.edu} 
           \and
           $^1$Department of Computer Science, College of Engineering, Wayne State University, Detroit, MI 48202\\
$^2$Division of Behavioral Health Sciences, Department of Family Medicine and Public Health Sciences, Wayne State University School of Medicine, Detroit, MI 48202\\
$^3$Director, Center for Translational Behavioral Research, Department of Behavioral Sciences and Social Medicine, Florida State University, FL 32306\\  
$^4$Department of Oncology, Wayne State University/Karmanos Cancer Institute, Detroit, MI 48201\\
$^5$Department of Public Health Sciences, Henry Ford Health System, Detroit, MI 48202\\
$^6$Department of Humanities, Health and Society, Wertheim College of Medicine, Florida International University, Miami, FL 33199    
}

@@

\keywords{sequential analysis \and hidden markov model \and closed frequent pattern mining \and motivational interviewing \and weight loss \and adolescents}
\end{abstract}

\section{Introduction}
\label{sec:intro}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change \cite{miller2013motivational}. The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills \cite{miller2013motivational}. The relational hypothesis suggests that counselors’ use of accurate empathy, positive regard, and congruence create the “spirit of MI”, an optimal therapeutic state to explore behavior change. MI’s technical hypothesis \cite{miller2009toward} states that counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient “change talk”. Change talk are patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change. Previous studies \cite{apodaca2009mechanisms} have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention \cite{walker2011influence}. In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk). Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk \cite{magill2014technical}. These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers’ ability to test MI’s technical hypothesis.

@@

\begin{table}
\caption{MYSCOPE codebook.}
\label{tab:codebook}   
\begin{tabularx}{\textwidth}{lll}
\hline\noalign{\smallskip}
first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number \\
number & number & number \\
\noalign{\smallskip}\hline
\end{tabularx}
\end{table}

@@

\begin{figure*}
\includegraphics[width=1.0\textwidth]{figures/bic-vs-hidden-states.eps}
\caption{Bayesian information criterion (BIC) of HMM models of successful (left) and unsuccessful (right) interviews by varying the number of hidden states.}
\label{fig:bic}
\end{figure*}

@@

\begin{figure*}
\includegraphics[width=0.95\textwidth]{figures/fpm.eps}
\caption{Different type of itemsets is illustrated with a frequent pattern mining method applied on transaction database by considering minimum support count 2, where TID represents the transaction ID of the corresponding items.}
\label{fig:fpm}
\end{figure*}

@@

\begin{sidewaystable}
\vspace{15\baselineskip}

\caption{Hidden Markov Model Emission Matrices}
\label{tab:emission}
\begin{tabularx}{\textwidth}{lllllllll}
\hline\noalign{\smallskip}
first & second & third & first & second & third & first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number & number & number & number & number & number & number\\
number & number & number & number & number & number & number & number & number\\
\noalign{\smallskip}\hline
\end{tabularx}

@@

\caption{Hidden Markov Model Transition Matrices}
\label{tab:transition}  
\begin{tabularx}{\textwidth}{lll}
\hline\noalign{\smallskip}
first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number \\
number & number & number \\
\noalign{\smallskip}\hline
\end{tabularx}
\end{sidewaystable}

@@

\begin{table}
\begin{threeparttable}
\caption{Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences.}
\label{tab:patterns}   
\begin{tabularx}{\textwidth}{lll}
\hline\noalign{\smallskip}
first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number number number number number number \\
number number number number & number number number & number number  number \\
\noalign{\smallskip}\hline
\end{tabularx}
\begin{tablenotes}
      \small
      \item This is where authors provide additional information about the data, including whatever notes are needed.
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

@@

\bibliographystyle{natbib}       % JHIR format
\bibliography{references}   % name your BibTeX data base

\end{document}
